---
title: "Practicum 2"
output:
  html_document:
    toc: true
---

```{r}
library(sqldf)
library(RSQLite)
library(XML)
```

```{r}
fpath = "~/Google\ Drive/Northeastern\ University/CS5200/Practicum\ 2/"
dbfile = "practicum2.db"

# if database file already exists, we connect to it, otherwise
# we create a new database
dbcon <- dbConnect(RSQLite::SQLite(), paste0(fpath,dbfile))
```

# Part 1 
## Part 1 Q1
Assumption: Each article belongs to one journal.

## Part 1 Q2
Create database Statements

```{sql connection=dbcon}
DROP TABLE IF EXISTS AuthorArticle;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS Author;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS History;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS Article;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS Journal;
```

```{sql connection=dbcon}
CREATE TABLE Author (
  author_id INTEGER PRIMARY KEY,
  fore_name TEXT NOT NULL,
  last_name TEXT NOT NULL,
  initials TEXT,
  affiliation TEXT
);
```

```{sql connection=dbcon}
CREATE TABLE Journal (
  journal_id INTEGER PRIMARY KEY,
  title TEXT,
  issn TEXT,
  pub_date DATE,
  volume INTEGER,
  issue INTEGER
);
```

```{sql connection=dbcon}
CREATE TABLE Article (
  article_id INTEGER PRIMARY KEY,
  title TEXT,
  language TEXT,
  journal_id INTEGER,
  FOREIGN KEY (journal_id) REFERENCES Journal(journal_id)
);
```


```{sql connection=dbcon}
CREATE TABLE History (
  article_id INTEGER,
  pub_status TEXT,
  history_date DATE,
  PRIMARY KEY (article_id, pub_status),
  FOREIGN KEY (article_id) REFERENCES Article(article_id)
);
```

```{sql connection=dbcon}
CREATE TABLE AuthorArticle (
  author_id INTEGER,
  article_id INTEGER,
  PRIMARY KEY (author_id, article_id),
  FOREIGN KEY (author_id) REFERENCES Author(author_id),
  FOREIGN KEY (article_id) REFERENCES Article(article_id)
)
```


# Part 1 Q3

## XML connection
Establishes a connection to the XML file.
```{r}
path = "~/Google\ Drive/Northeastern\ University/CS5200/Practicum\ 2/"
xmlFile <- "pubmed_sample.xml"
xmlDOM <- xmlParse(file = paste0(path, xmlFile), validate = F)
r <- xmlRoot(xmlDOM)
# df <- xmlToDataFrame(paste0(path,xmlFile), stringsAsFactors = F)
```

## Functions
Here we create functions that can be used to parse things that go into a particular dataframe or help separate out code. 

Checks if a row exists in a dataframe.
```{r rowExists}
rowExists <- function (aRow, aDF)
{
  # check if that address is already in the data frame
  n <- nrow(aDF)
  c <- ncol(aDF)
  
  if (n == 0)
  {
    # data frame is empty, so can't exist
    return(0)
  }
  
  for (a in 1:n)
  {
    # check if all columns match for a row; ignore the aID column
    if (all(aDF[a,] == aRow[1,]))
    {
      # found a match; return it's ID
      return(a)
    }
  }
  
  # none matched
  return(0)
}
```

Parses a single date node that is a child of history.
```{r parseHistory}
parseHistory <- function (aHistoryNode)
{
  attributes <- xmlAttrs(aHistoryNode)
  pub_status <- attributes[1]
  year <- xpathSApply(aHistoryNode, "./Year", xmlValue)
  month <- xpathSApply(aHistoryNode, "./Month", xmlValue)
  day <- xpathSApply(aHistoryNode, "./Day", xmlValue)
  
  date <- paste(year, month, sep ="/")
  date <- paste(date, day, sep= "/")
  
  #Converts date to correct format.
  date <- as.Date(date, "%Y/%m/%d")
  
  #Changes it back to a character to insert into a dataframe.
  date <- as.character(date)
  newHistory.df <- data.frame(pub_status, date,
                              stringsAsFactors = F)
  return(newHistory.df)
}
```

Parses an author node and returns a dataframe with an authors last name, fore name and affiliation.
```{r parseAuthor}
parseAuthor <- function (anAuthorNode)
{
  last_name <- xpathSApply(anAuthorNode, "./LastName", xmlValue)
  first_name <- xpathSApply(anAuthorNode, "./ForeName", xmlValue)
  initials <- xpathSApply(anAuthorNode, "./Initials", xmlValue)
  affiliation <- xpathSApply(anAuthorNode, "./Affiliation", xmlValue)
  if (length(affiliation) == 0) {
    affiliation <- "NA"
    #String NA is needed to use row exist comparison.
    newAuthor.df <- data.frame(first_name, last_name, initials, affiliation,
                             stringsAsFactors = F)
  }
  else {
  newAuthor.df <- data.frame(first_name, last_name, initials, affiliation,
                             stringsAsFactors = F)
  }
  return(newAuthor.df)
}
```

Parses the journal node.
```{r parseJournal}
parseJournal <- function (aJournalNode)
{
  issn <- xpathSApply(aJournalNode, "./ISSN", xmlValue)
  date <- xpathSApply(aJournalNode, "./JournalIssue/PubDate/MedlineDate", xmlValue)

  #Not a medline date
  if(length(date) == 0){
    year <- xpathSApply(aJournalNode, "./JournalIssue/PubDate/Year", xmlValue)
    month <- xpathSApply(aJournalNode, "./JournalIssue/PubDate/Month", xmlValue)
    # If no month for medline, assume month is January.
    if (length(month) == 0) {
      month <- "Jan"
    }
    day <-xpathSApply(aJournalNode, "./JournalIssue/PubDate/Day", xmlValue)
    
    # If there is no day for medline, assume first day of the month.
    if (length(day) == 0) {
      day <- "01"
    }
    
    date <- paste(year, month, sep ="/")
    date <- paste(date, day, sep= "/")
  } else {
    #medline date has months with a dash between them
    #Medline dates also do not have days.
    split_date <- strsplit(date, "-")
    #Remove the month after the dash
    date <- split_date[[1]][1]
    split_date <- strsplit(date, " ")
    date <- paste(split_date[[1]][1], split_date[[1]][2], sep = "/")
    # Cardinal date value
    # Since two months are listed, set day to 15
    date <- paste(date, "15", sep = "/")
  }
  
  #Turn the date format into the R date format.
  date <- as.Date(date, "%Y/%b/%d")
  #Change it back to a character so that it writes to the dataframe correctly.
  date <- as.character(date)
  
  journal_title <- xpathSApply(aJournalNode, "./Title", xmlValue)
  volume <- xpathSApply(aJournalNode, "./JournalIssue/Volume", xmlValue)
  issue <- xpathSApply(aJournalNode, "./JournalIssue/Issue", xmlValue)
  newJournal.df <- data.frame(journal_title, issn, date, volume, issue,
                              stringsAsFactors = F)
  
  return(newJournal.df)
}
```



## Dataframes
Create the dataframes that will be tables in the database.

This counts the number of articles so that I know how many loops I need to go through and also pre load the space for the data in the dataframe.
```{r}
#Can count either PubmedArticle or Article by itself.
numArticles <- xpathSApply(r, "count(//PubmedArticle)", xmlValue)

numArticles
```

```{r}
article_df <- data.frame(article_id = vector(mode = "integer", length = numArticles),
                         title = vector (mode = "character", length = numArticles),
                         language = vector (mode = "character", length = numArticles),
                         journal_id = vector (mode = "integer", length = numArticles),
                         stringsAsFactors = F
                         )
```

```{r}
journal_df <- data.frame(journal_id = integer(),
                         title = character(),
                         issn = character(),
                         pub_date = character(),
                         volume = integer(),
                         issue = integer(),
                         stringsAsFactors = F
                         )
```

```{r}
author_df <- data.frame(author_id = integer(),
                        fore_name = character(),
                        last_name = character(),
                        initials = character(),
                        affiliation = character(),
                        stringsAsFactors = F)
```

```{r}
history_df <- data.frame(article_id = integer(),
                         pub_status = character(),
                         history_date = character(),
                         stringsAsFactors = F)
```


```{r}
author_article_df <- data.frame(article_id = integer(),
                             author_id = integer(),
                             stringsAsFactors = F)
                             
```




## Main loop for node traversal
Some assumptions:
Author's first name, last name and initial uniquely identify them, and affiliation does not matter for identification purposes.


```{r}
for (i in 1:numArticles)
{
  pubmed_article <- (r[[i]])
  #Extract the article node with an xpath expression.
  article <- xpathSApply(pubmed_article, "./MedlineCitation/Article")
  
  
  title <- xpathSApply(article[[1]], "./ArticleTitle", xmlValue)
  language <- xpathSApply(article[[1]], "./Language", xmlValue)

  article_df$article_id[i] <- i
  article_df$title[i] <- title
  article_df$language[i] <- language
  
  #Extract the journal node.
  journal <- xpathSApply(article[[1]], "./Journal")
  newJournal <- parseJournal(journal[[1]])
  pk.journal <- rowExists(newJournal, journal_df[,2:ncol(journal_df)])
  
  # Checks if journal has been added before.
  if (pk.journal == 0){  
    n_journal <- nrow(journal_df) + 1
    journal_df[n_journal, 2:ncol(journal_df)] <- newJournal[1,]
    journal_df$journal_id[n_journal] <- n_journal
    pk.journal <- n_journal
  }
  article_df$journal_id[i] <- pk.journal

  #Gets the list of authors by retrieving the AuthorList node.
  author_list <- xpathSApply(article[[1]], "./AuthorList")
  
  #Counts the number of authors in this article node
  numAuthors <- xpathSApply(author_list[[1]], "count(./Author)", xmlValue)
  
  #Goes through each author in the author list and extracts the information.
  for (x in 1:numAuthors)
  {
    #This points to a single author node.
    authorPointer <- author_list[[1]][x]
    
    #Pass the author pointer to the parse function
    newAuthor <- parseAuthor(authorPointer[[1]])
    
    n_author_columns <- ncol(author_df) - 1
    pk.author <- rowExists(newAuthor[,1:ncol(newAuthor) - 1], author_df[,2:n_author_columns])
    #pk.author <- rowExists(newAuthor, author_df[,2:ncol(author_df)])

    #If author not in the list.
    if (pk.author == 0) {
      n_author <- nrow(author_df) + 1
      author_df[n_author, 2:ncol(author_df)] <- newAuthor[1,]
      author_df$author_id[n_author] <- n_author
      pk.author <- n_author
    }
    # If author's affiliation is NA, add the new author's node affiliation.
    if (newAuthor$affiliation[1] != "NA") {
      author_df$affiliation[pk.author] <- newAuthor$affiliation[1]
    }
    #Puts the author and article together in a dataframe.
    numAuthorBook <- nrow(author_article_df) + 1
    author_article_df[numAuthorBook, 1] <- i
    author_article_df[numAuthorBook, 2] <- pk.author
    #author_article_df$article_id[numAuthorBook] <- 1
    #author_article_df$author_id[numAuthorBook] = pk.author
    
  }
  
  #History portion
  #Extracts the history node.
  history_list <- xpathSApply(pubmed_article, "./PubmedData/History")
  
  #Counts the number of history nodes.
  numDates <-xpathSApply(history_list[[1]], "count(PubMedPubDate)", xmlValue)
  for (y in 1:numDates) {
    pubmedDatePointer <- history_list[[1]][y]
    newDate <- parseHistory(pubmedDatePointer[[1]])
    
    n_history <- nrow(history_df) + 1
    
    history_df[n_history, 2:ncol(history_df)] <- newDate[1,]
    history_df$article_id[n_history] <- i
  }
  
  
}
```

## Writing to database
```{r}
dbWriteTable(dbcon, "Author", author_df, append = TRUE, header = TRUE, row.names=FALSE)
```

```{r}
dbWriteTable(dbcon, "Journal", journal_df, append = TRUE, header = TRUE, row.names=FALSE)
```

```{r}
dbWriteTable(dbcon, "Article", article_df, append = TRUE, header = TRUE, row.names=FALSE)
```

```{r}
dbWriteTable(dbcon, "History", history_df, append = TRUE, header = TRUE, row.names=FALSE)
```

```{r}
dbWriteTable(dbcon, "AuthorArticle", author_article_df, append = TRUE, header = TRUE, row.names=FALSE)
```


```{r}
validation <-sqldf("Select article_df.title, article_df.article_id, author_df.fore_name, author_df.last_name FROM article_df
        INNER JOIN author_article_df ON article_df.article_id = author_article_df.article_id
        INNER JOIN author_df ON author_article_df.author_id = author_df.author_id")
```



```{r}
dbDisconnect(dbcon)
```
